Comparing MCMC algorithms in Stochastic Volatility Models using Simulation Based Calibration

### 2023-10-07 manuscript/main.tex
Updates from Overleaf

    To illustrate this problem, the stochastic volatility model is estimated with two different MCMC algorithms on real data. The data are the de-meaned log returns of the 2023 January-September S\&P 500 Index. Below are the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ and corresponding summary statistics.

    \begin{figure}[h]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{Posterior samples from two MCMC algorithms using S\&P 500 data. While the shape of the distributions are similar, the modes of method 2 is much higher and the tails of method 1 are fatter for $\phi$ and $\sigma^2$. Convergence diagnostics do not tell us which estimate is "more correct" since we don't observe the true model or true data generating process.}
    \end{figure}

    These estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$  has a fatter left tail for and $\sigma^2$ has heavier right tails for sampling method 1. Furthermore, sampling method 2 has higher peaks at their mode relative to method 1 with a tighter spread. Based on this information alone, there is no certainty in whether method 1 or method 2 is has a more correct estimate conditional on data and model. There are other methods for model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

@@

    Additionally, HMC struggles to generate efficient posterior samples from the centered parameterisation of the stochastic volatility model. Figure \ref{fig:hmcess} visualises the distribution of ESS for the static parameters after 5000 SBC iterations. The non centered model appears to generate independent samples much more efficiently than the centered model. This is consistent with HMC struggliing to sample overall from the posteriors of $\sigma^2$ and $\phi$.

@@

@@

    The ESS estimates for this model indicate a high degree of autocorrelation and difficulty in generating independent samples from both parameterisations. The numerical results are summarised in table x. The estimate for $\mu$ in the reparameterised model exhibits a long right tail (longer than the 10,000 posterior draws suggesting the presence of negatively autocorrelated draws). Overall, the bespoke MCMC is quite insufficient at sampling the static parameters.

@@

@@

>>>>

    To illustrate this problem, the stochastic volatility model is estimated with two different MCMC algorithms on real data. The data are the de-meaned log returns of the 2023 January-September S\&P 500 Index. Figure \ref{fig:realdataex} visualises the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ with corresponding summary statistics in Tables \ref{tab:realdata}.

    \begin{figure}[h]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{Posterior samples from the Hamiltonian Monte Carlo (HMC) and Kim Sherphard and Chib (KSC) MCMC samplers using S\&P 500 data. While the shape of the distributions are similar, the modes of method 2 is much higher and the tails of method 1 are fatter for $\phi$ and $\sigma^2$. Convergence diagnostics do not tell us which estimate is "more correct" since we don't observe the true model or true data generating process.}
        \label{fig:realdataex}
    \end{figure}

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter&  MCMC&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  KSC&-2.03 & -0.472 & -0.354 & -0.355 & -0.231 & 1.31 \\
     $\mu$&  HMC&-4.55 & -0.508 & -0.373 & -0.370 & -0.229 &2.86  \\\hline 
     $\phi$&  KSC&0.384 & 0.875 & 0.928 & 0.902 & 0.958 & 0.997 \\
     $\phi$&  HMC&0.438 & 0.910 & 0.953 & 0.929 & 0.977 &1.00  \\ \hline 
     $\sigma^2$&  KSC&0.00237 & 0.00891 & 0.0138 & 0.0178 & 0.0210 & 0.130 \\ 
     $\sigma^2$&  HMC&0.00209 & 0.00731 & 0.0105 & 0.0131 & 0.0159 &0.107 \\ \hline
        \end{tabular}
        \caption{Summary statistics for KSC sampler}
        \label{tab:realdata}
    \end{table}

    These estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$  has a fatter left tail for the KSC sampler and $\sigma^2$ has heavier right tails for HMC. This is evident in the quantiles with the 25th quantile for $\phi$ being smaller at 0.875 for KSC compared to 0.910 in HMC. Furthermore, sampling method 2 has higher peaks at their mode relative to method 1 with a tighter spread. Based on this information alone, there is no certainty in whether method 1 or method 2 is has a more correct estimate conditional on data and model. There are other methods for model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

@@

    Additionally, HMC struggles to generate efficient posterior samples from the centered parameterisation of the stochastic volatility model. Figure \ref{fig:hmcess} and Table \ref{tab:hmcess} summarises the distribution of ESS for the static parameters after 5000 SBC iterations. The non centered model appears to generate independent samples much more efficiently than the centered model. This is consistent with HMC struggliing to sample overall from the posteriors of $\sigma^2$ and $\phi$.

@@

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter &  Type&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  Centered&3.79 & 99.7 & 184. & 305. & 457. & 1997.  \\
     $\mu$&  Reparam&14.5 & 738. & 1099. & 1143. & 1500. & 2993.  \\\hline 
     $\phi$&  Centered&1.44 & 11.0 & 21.7 & 26.3 & 34.1 & 635.  \\
     $\phi$&  Reparam&48.8 & 355. & 476. & 521. & 651. & 1445.   \\ \hline 
     $\sigma^2$&  Centered& 1.17 & 2.99 & 7.04 & 8.05 & 11.3 & 90.7 \\ 
     $\sigma^2$&  Reparam&46.1 & 403. & 553. & 585. & 734. & 1486. \\ \hline
        \end{tabular}
        \caption{HMC: ESS for centered and reparameterised stochastic volatility model.}
        \label{tab:hmcess}
    \end{table}

@@

    The ESS estimates for this model indicate a high degree of autocorrelation and difficulty in generating independent samples from both parameterisations. The numerical results are summarised in Figure \ref{kscess} and Table \ref{tab:kscess}. The estimate for $\mu$ in the reparameterised model exhibits a long right tail (longer than the 10,000 posterior draws suggesting the presence of negatively autocorrelated draws). Overall, the bespoke MCMC is quite insufficient at sampling the static parameters.

@@

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter &  Type&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  Centered&9.57 & 352. & 559. & 1482. & 1552. & 9907.\\
     $\mu$&  Reparam&1.99 & 205. & 569. & 4174. & 6384. & 34055.\\\hline 
     $\phi$&  Centered&2.47 & 74.0 & 103. & 112. & 130. & 1222.\\
     $\phi$&  Reparam&1.03 & 83.4 & 132. & 325. & 419. & 1907. \\ \hline 
     $\sigma^2$&  Centered&2.33 & 29.0 & 39.5 & 44.2 & 52.8 & 384. \\ 
     $\sigma^2$&  Reparam&1.25 & 47.4 & 73.4 & 399. & 591. & 6618. \\ \hline
        \end{tabular}
        \caption{KSC: ESS for centered and reparameterised stochastic volatility model.}
        \label{tab:kscess}
    \end{table}

@@

\begin{table}[H]
    \centering
    \begin{tabular}{lccc} 
          $\omega$ &$Pr(\omega = i)$&  $m_i$&  $\nu^2_i$\\ 
          1&0.00730  &  -10.12999&  5.79596\\ 
          2&0.10556  &   -3.97281 &  2.61369\\ 
          3&0.00002 &  -8.56686 &   5.17950\\ 
          4&0.04395 &  2.77786  &   0.16735 \\ 
          5&0.34001&   0.61942    &  0.64009\\ 
          6&0.24566 &  1.79518    &  0.34023 \\ 
          7&0.25750 &  -1.08819    &  1.26261\\ 
    \end{tabular} 
\end{table}
